---
title: "BilboardHot100Ones"
author: "Mateusz Nowaczek, Gabriel Nowak, Jan Pawelec"
date: "2026-05-01"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

# What about

The project is about analyzing a "Billboard Hot 100 Number Ones" dataset
from tidyverse. We have chosen this particular set, because everybody
love music.

Source:
<https://docs.google.com/spreadsheets/d/1j1AUgtMnjpFTz54UdXgCKZ1i4bNxFjf01ImJ-BqBEt0/edit?gid=1974823090#gid=1974823090>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

library(tidyverse)
library(tidytuesdayR)
library(janitor)
```

# 1. Data Cleaning and Preparation

```{r}
tuesdata <- tidytuesdayR::tt_load('2025-08-26')

billboard_raw <- tuesdata$billboard 
```

Let's have a look on our data.

```{r}
glimpse(billboard_raw)
```

## 1.1. Is na?

```{r}
colSums(is.na(billboard_raw))
```

## 1.2. Complex cleaning

Now we can clean missing values and extra information we have no need to
use.

```{r}
billboard_clean <- billboard_raw %>%
  clean_names() %>%
  select(-contains("lyrics"), -contains("text")) %>%
  mutate(
    across(where(is.character), str_squish)
  ) %>%
  mutate(
    date = lubridate::ymd(date)
  ) %>%
  drop_na(artist, song, date) %>%
  distinct()
  
colSums(is.na(billboard_clean))
```

## 1.3 Preparation of new columns
Ten new columns for extra insight.

```{R}
billboard_test <- billboard_clean

billboard_test <- billboard_test %>% mutate(decade = paste0(floor(year(date) / 10) * 10 , "s"))
billboard_test <- billboard_test %>% mutate(year = paste0(floor(year(date))))
# First two allow for better analysis of the time of release
billboard_test <- billboard_test %>% mutate(is_collab = str_detect(artist, "feat\\.|&|,|and"))
billboard_test <- billboard_test %>% mutate(artist_count = str_count(artist, "&|,|feat\\.") + 1)
# These two make it easier to see if songs performed by multiple people are more popular
billboard_test <- billboard_test %>% mutate(song_title_length = nchar(song))
# This one can be used to see if longer song titles may lead to lower popularity

billboard_test <- billboard_test %>%
  mutate(
  season = case_when(
    month(date) %in% c(12,1,2) ~ "Winter",
    month(date) %in% c(3,4,5) ~ "Spring",
    month(date) %in% c(6,7,8) ~ "Summer",
    TRUE ~ "Autumn"
  )
)
## This allows us to see in what season does the most hit songs get produced
billboard_test %>% count(season)

billboard_test <- billboard_test %>% 
  mutate(
    is_solo = if_else(artist_count == 1, TRUE, FALSE),
    
    month = month(date, label = TRUE),
    
    artist_name_length = nchar(artist),
    
    title_word_count = str_count(song, " ") + 1
  )
```

Let's out set after changes.
```{r}
glimpse(billboard_test)
```

# 2. Exploratory Data Analysis

## 2.1 Interesting statistics

Length of the song title. We can see that there are longer song titles
(max == 66) but they tend to be much shorter (mean == 16) [characters
long]

```{R}
max(billboard_test$song_title_length)
min(billboard_test$song_title_length)
mean(billboard_test$song_title_length)
```

Percent of Artist whom wrote the songs only themselves.

```{R}
(sum(billboard_test["artist_is_only_songwriter"]) / count(billboard_test["artist_is_only_songwriter"])) * 100
```

And on the other hand, percent of artist whom haven't even contributed
to the writing process.

```{R}
(count(billboard_test["artist_is_a_songwriter"]) - sum(billboard_test["artist_is_a_songwriter"]))/count(billboard_test["artist_is_a_songwriter"]) *100
```

Percentage of songs that are collabs.

```{R}
sum(billboard_test$is_collab) / count(billboard_test) * 100
```

## 2.2 Data visualization

This chart shows us that it doesn't matter when the song was released
season wise

```{R}
ggplot(billboard_test, aes(x = season, fill = season)) +
  geom_bar()

```

This one shows that most popular songs were written in the 70s and 80s. Or they last shorter on no. 1 position.
```{R}
ggplot(billboard_test, aes(x = decade, fill = decade)) +
  geom_bar()
```

```{R}
# number of hits per year
billboard_test %>%
  mutate(year = as.numeric(year)) %>%
  count(year) %>%
  ggplot(aes(x = year, y = n)) +

  geom_line(
    color = "#4E79A7",
    linewidth = 1.2
  ) +

  geom_point(
    color = "#4E79A7",
    size = 2
  ) +

  scale_x_continuous(
    breaks = seq(1955, 2025, by = 5)
  ) +

  labs(
    title = "Number of Billboard #1 Hits per Year",
    subtitle = "Distribution of songs across years",
    x = "Year",
    y = "Number of hits"
  ) +

  theme_minimal(base_size = 13) +

  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

```{R}
# Popularity of collab songs
billboard_test %>%
  mutate(year = as.numeric(year)) %>%
  group_by(year) %>%
  summarise(
    collab_percent = mean(is_collab, na.rm = TRUE) * 100
  ) %>%
  ggplot(aes(x = year, y = collab_percent)) +
  geom_line(color = "#4E79A7", linewidth = 1.2) +
  geom_point(color = "#4E79A7", size = 2) +
  theme_minimal(base_size = 13) +

  labs(
    title = "Popularity of Collaboration Songs Over Time",
    subtitle = "Percentage of Billboard #1 songs that were collaborations",
    x = "Year",
    y = "% of collaboration songs"
  )
```

```{R}
ggplot(billboard_test, aes(x = song_title_length)) +
  geom_histogram(
    binwidth = 2,
    fill = "#4E79A7",
    color = "white",
    alpha = 0.9
  ) +
  labs(
    title = "Distribution of Song Title Lengths",
    subtitle = "Most Billboard #1 songs have relatively short titles",
    x = "Song title length (characters)",
    y = "Number of songs"
  ) +
  theme_minimal(base_size = 13)
```

```{R}
# Titles got slightly shorter over time
ggplot(billboard_test,
       aes(x = decade, y = song_title_length, fill = decade)) +
  geom_boxplot()
```

```{R}
# Top 10 artists with most hit songs
billboard_test %>%
  count(artist, sort = TRUE) %>%
  slice_head(n = 10) %>%
  ggplot(aes(x = reorder(artist, n), y = n)) +
  geom_col(fill="#4E79A7") +
  coord_flip() +
  labs(title="Top 10 aritsts with most hit songs")
```

```{R}
billboard_test %>%
  count(decade, season) %>%

  ggplot(aes(x = decade, y = season, fill = n)) +

  geom_tile(color = "white", linewidth = 1) +

  geom_text(
    aes(label = n),
    color = "white",
    size = 5
  ) +

  scale_fill_gradient(
    low = "#A8DADC",
    high = "#1D3557"
  ) +

  labs(
    title = "Billboard #1 Hits by Season and Decade",
    subtitle = "Darker colors indicate more songs",
    x = "Decade",
    y = "Season",
    fill = "Songs"
  ) +

  theme_minimal(base_size = 13) +

  theme(
    panel.grid = element_blank(),
    axis.text.x = element_text(face = "bold"),
    axis.text.y = element_text(face = "bold")
  )
```
```{R}
ggplot(billboard_test, aes(x = decade, y = title_word_count, fill = decade)) +
  geom_boxplot(alpha = 0.7) +
  labs(
    title = "Word Count in Song Titles by Decade",
    x = "Decade",
    y = "Number of words in title"
  ) +
  theme_minimal()
```

Solo vs colab

```{R}
billboard_test %>%
  count(is_solo) %>%
  ggplot(aes(x = is_solo, y = n, fill = is_solo)) +
  geom_col() +
  labs(
    title = "Solo Artists vs Collaborations (#1 Hits)",
    x = "Is it a solo artist?",
    y = "Number of songs"
  ) +
  theme_minimal()
```


## 2.3 Data visualization for model

```{r}
billboard_test %>%
  group_by(decade) %>%
  summarize(pct_black = mean(artist_black, na.rm = TRUE) * 100) %>%
  ggplot(aes(x = decade, y = pct_black, group = 1)) +
  geom_line(color = "#E15759", linewidth = 1.2) +
  geom_point(color = "#E15759", size = 3) +
  labs(
    title = "Percentage of #1 Hits by Black Artists Over Time",
    x = "Decade",
    y = "% of Hits by Black Artists"
  ) +
  theme_minimal()
```

Filter to keep only clear Male (1) or Female (2) solo artists.
```{R}
billboard_test %>%
  filter(artist_male %in% c(1, 2)) %>%
  mutate(Gender = ifelse(artist_male == 1, "Male", "Female")) %>%
  ggplot(aes(x = Gender, y = danceability, fill = Gender)) +
  geom_boxplot(alpha = 0.7) +
  scale_fill_manual(values = c("#F28E2B", "#4E79A7")) +
  labs(
    title = "Song Danceability: Male vs Female Artists",
    x = "Artist Gender",
    y = "Danceability Score"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

```{r}
billboard_test %>%
  # Wybieramy tylko solowych artystów (1 = Mężczyzna, 2 = Kobieta)
  filter(artist_male %in% c(1, 2)) %>%
  mutate(Gender = ifelse(artist_male == 1, "Male", "Female")) %>%
  count(decade, Gender) %>%
  ggplot(aes(x = decade, y = n, color = Gender, group = Gender)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  scale_color_manual(values = c("Male" = "#4E79A7", "Female" = "#F28E2B")) +
  labs(
    title = "Number of #1 Hits: Male vs Female Artists Over Time",
    x = "Decade",
    y = "Number of Hit Songs",
    color = "Artist Gender"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

```{r}
billboard_test %>%
  mutate(Race = ifelse(artist_black == 1, "Black Artists", "Other Artists")) %>%
  ggplot(aes(x = Race, y = danceability, fill = Race)) +
  geom_boxplot(alpha = 0.8, color = "darkgray") +
  scale_fill_manual(values = c("#E15759", "#76B7B2")) +
  labs(
    title = "Song Danceability by Artist Race",
    x = "Artist Demographic",
    y = "Danceability Score"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none")
```

```{r}
billboard_test %>%
  # Filter only clear Male (1) or Female (2) solo artists
  filter(artist_male %in% c(1, 2)) %>%
  mutate(Gender = ifelse(artist_male == 1, "Male", "Female")) %>%
  ggplot(aes(x = acousticness, fill = Gender)) +
  geom_density(alpha = 0.6) +
  scale_fill_manual(values = c("Male" = "#4E79A7", "Female" = "#F28E2B")) +
  labs(
    title = "Acousticness of #1 Hits: Male vs Female Artists",
    x = "Acousticness Score (0 = Electronic/Produced, 100 = Acoustic)",
    y = "Density",
    fill = "Artist Gender"
  ) +
  theme_minimal(base_size = 13)
```

# 3. Modeling
```{r include=FALSE}
library(tidymodels)
library(kknn)

model_data <- billboard_test %>%
  mutate(
    year_num = as.numeric(year),
    is_black = factor(ifelse(artist_black == 1, "yes", "no"), levels = c("yes", "no"))
  ) %>%
  drop_na(year_num, danceability, energy, acousticness, is_black)

set.seed(67) 
data_split <- initial_split(model_data, prop = 0.75, strata = is_black)
train_data <- training(data_split)
test_data  <- testing(data_split)

cv_folds <- vfold_cv(train_data, v = 5, strata = is_black)
```

## 3.1 Research Question and Hypothesis

Can we predict the demographic profile (gender and race) of a no.1 hit artist based on the song's audio features (danceability, acousticness, energy) and the year of release? If a song becomes a hit today, what kind of artist is most likely behind it?

Hypothesis: Based on our Exploratory Data Analysis, we hypothesize that highly danceable songs released in more recent decades are strongly associated with Black artists, due to the rise of R&B and Hip-Hop genres.

Let's also remember about normalization for knn.
```{r}
knn_recipe <- recipe(is_black ~ year_num + danceability + energy + acousticness, data = train_data) %>%
  step_normalize(all_numeric_predictors())
```

## 3.2 KNN

```{R}
knn_spec <- nearest_neighbor(neighbors = tune()) %>%
  set_engine("kknn") %>%
  set_mode("classification")

knn_wf <- workflow() %>%
  add_recipe(knn_recipe) %>%
  add_model(knn_spec)

k_grid <- tibble(neighbors = c(5, 10, 20, 30, 50, 75, 100))

set.seed(314)
knn_tuning <- tune_grid(
  knn_wf,
  resamples = cv_folds,
  grid = k_grid,
  metrics = metric_set(roc_auc, accuracy)
)

autoplot(knn_tuning) +
  labs(title = "KNN Tuning: Performance vs. Number of Neighbors (k)",
       subtitle = "Finding the optimal 'k' using 5-fold cross-validation") +
  theme_light()
```

```{R}
best_k <- select_best(knn_tuning, metric = "roc_auc")

final_knn_wf <- finalize_workflow(knn_wf, best_k)
knn_last_fit <- last_fit(final_knn_wf, split = data_split)


knn_metrics <- collect_metrics(knn_last_fit)
knn_metrics
```

## 3.3 Logistic Regression
```{R}
log_spec <- logistic_reg() %>% set_engine("glm")
log_wf <- workflow() %>% add_recipe(knn_recipe) %>% add_model(log_spec)

log_last_fit <- last_fit(log_wf, split = data_split)

fitted_log <- fit(log_wf, data = train_data)
tidy(fitted_log)
```

## 3.4 Model Evaluation and Comparison

```{R}
knn_predictions <- collect_predictions(knn_last_fit)

conf_mat_obj <- conf_mat(knn_predictions, truth = is_black, estimate = .pred_class)
autoplot(conf_mat_obj, type = "heatmap") +
  labs(title = "KNN Confusion Matrix (Test Data)")
```
```{R}
roc_obj <- roc_curve(knn_predictions, truth = is_black, .pred_yes)
autoplot(roc_obj) +
  labs(title = "KNN ROC Curve",
       subtitle = paste("Test Set AUC =", round(knn_metrics$.estimate[2], 3))) +
  theme_light()
```

```{r}
knn_preds <- collect_predictions(knn_last_fit) %>% mutate(model = "KNN")
log_preds <- collect_predictions(log_last_fit) %>% mutate(model = "Logistic Regression")

combined_preds <- bind_rows(knn_preds, log_preds)

combined_preds %>%
  group_by(model) %>%
  roc_curve(truth = is_black, .pred_yes) %>%
  autoplot() +
  labs(
    title = "ROC Curve Comparison: KNN vs Logistic Regression",
    subtitle = "Comparing the predictive power of both models",
    color = "Model"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom")

knn_metrics <- collect_metrics(knn_last_fit) %>% mutate(model = "KNN")
log_metrics <- collect_metrics(log_last_fit) %>% mutate(model = "Logistic Regression")

bind_rows(knn_metrics, log_metrics) %>%
  select(model, .metric, .estimate) %>%
  filter(.metric %in% c("accuracy", "roc_auc")) %>%
  pivot_wider(names_from = .metric, values_from = .estimate) %>%
  rename(Model = model, Accuracy = accuracy, `ROC AUC` = roc_auc) %>%
  knitr::kable(digits = 3, caption = "Performance Metrics on Test Data")
```

# 4. Summary and Reflection

## 4.1 Project Summary and Key Findings

## 4.2 Limitations

## 4.3 Reflection
