---
title: "BilboardHot100Ones"
author: "Mateusz Nowaczek, Gabriel Nowak, Jan Pawelec"
date: "2026-05-01"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

# What about

The project is about analyzing a "Billboard Hot 100 Number Ones" dataset
from tidyverse. We have chosen this particular set, because everybody
love music.

Source:
<https://docs.google.com/spreadsheets/d/1j1AUgtMnjpFTz54UdXgCKZ1i4bNxFjf01ImJ-BqBEt0/edit?gid=1974823090#gid=1974823090>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

library(tidyverse)
library(tidytuesdayR)
library(janitor)
```

# 1. Data Cleaning and Preparation

```{r}
tuesdata <- tidytuesdayR::tt_load('2025-08-26')

billboard_raw <- tuesdata$billboard 
```

Let's have a look on our data.

```{r}
glimpse(billboard_raw)
```

## 1.1. Is na?

```{r}
colSums(is.na(billboard_raw))
```

## 1.2. Complex cleaning

Now we can clean missing values and extra information we have no need to
use.

```{r}
billboard_clean <- billboard_raw %>%
  clean_names() %>%
  select(-contains("lyrics"), -contains("text")) %>%
  mutate(
    across(where(is.character), str_squish)
  ) %>%
  mutate(
    date = lubridate::ymd(date)
  ) %>%
  drop_na(artist, song, date) %>%
  distinct()
  
colSums(is.na(billboard_clean))
```

```{R}
# a few new columns
billboard_test <- billboard_clean

billboard_test <- billboard_test %>% mutate(decade = paste0(floor(year(date) / 10) * 10 , "s"))
billboard_test <- billboard_test %>% mutate(year = paste0(floor(year(date))))



```



## This allows us to see in what season does the most hit songs get produced

```{R}
billboard_test <- billboard_test %>%
  mutate(
  season = case_when(
    month(date) %in% c(12,1,2) ~ "Winter",
    month(date) %in% c(3,4,5) ~ "Spring",
    month(date) %in% c(6,7,8) ~ "Summer",
    TRUE ~ "Autumn"
  )
)

billboard_test %>% count(season)
```
## we can now make a chart that highliths that comparison.

```{R}
ggplot(billboard_test, aes(x = season, fill = season)) +
  geom_bar()

```

```{R}
ggplot(billboard_test, aes(x = decade, fill = decade)) +
  geom_bar()
```

```{R}
(sum(billboard_test["artist_is_only_songwriter"]) / count(billboard_test["artist_is_only_songwriter"])) * 100
```

### This precent of Artist wrote the songs themselves

```{R}
(count(billboard_test["artist_is_songwriter"]) - sum(billboard_test["artist_is_songwriter"])
```