---
title: "BilboardHot100Ones"
author: "Mateusz Nowaczek, Gabriel Nowak, Jan Pawelec"
date: "2026-05-01"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

# What about

The project is about analyzing a "Billboard Hot 100 Number Ones" dataset
from tidyverse. We have chosen this particular set, because everybody
love music.

Source:
<https://docs.google.com/spreadsheets/d/1j1AUgtMnjpFTz54UdXgCKZ1i4bNxFjf01ImJ-BqBEt0/edit?gid=1974823090#gid=1974823090>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

library(tidyverse)
library(tidytuesdayR)
library(janitor)
```

# 1. Data Cleaning and Preparation

```{r}
tuesdata <- tidytuesdayR::tt_load('2025-08-26')

billboard_raw <- tuesdata$billboard 
```

Let's have a look on our data.

```{r}
glimpse(billboard_raw)
```

## 1.1. Is na?

```{r}
colSums(is.na(billboard_raw))
```

## 1.2. Complex cleaning

Now we can clean missing values and extra information we have no need to
use.

```{r}
billboard_clean <- billboard_raw %>%
  clean_names() %>%
  select(-contains("lyrics"), -contains("text")) %>%
  mutate(
    across(where(is.character), str_squish)
  ) %>%
  mutate(
    date = lubridate::ymd(date)
  ) %>%
  drop_na(artist, song, date) %>%
  distinct()
  
colSums(is.na(billboard_clean))
```

## 1.3 Preparation of new columns

Ten new columns for extra insight.

```{R}
billboard_test <- billboard_clean

billboard_test <- billboard_test %>% mutate(decade = paste0(floor(year(date) / 10) * 10 , "s"))
billboard_test <- billboard_test %>% mutate(year = paste0(floor(year(date))))
# First two allow for better analysis of the time of release
billboard_test <- billboard_test %>% mutate(is_collab = str_detect(artist, "feat\\.|&|,|and"))
billboard_test <- billboard_test %>% mutate(artist_count = str_count(artist, "&|,|feat\\.") + 1)
# These two make it easier to see if songs performed by multiple people are more popular
billboard_test <- billboard_test %>% mutate(song_title_length = nchar(song))
# This one can be used to see if longer song titles may lead to lower popularity

billboard_test <- billboard_test %>%
  mutate(
  season = case_when(
    month(date) %in% c(12,1,2) ~ "Winter",
    month(date) %in% c(3,4,5) ~ "Spring",
    month(date) %in% c(6,7,8) ~ "Summer",
    TRUE ~ "Autumn"
  )
)
## This allows us to see in what season does the most hit songs get produced
billboard_test %>% count(season)

billboard_test <- billboard_test %>% 
  mutate(
    is_solo = if_else(artist_count == 1, TRUE, FALSE),
    
    month = month(date, label = TRUE),
    
    artist_name_length = nchar(artist),
    
    title_word_count = str_count(song, " ") + 1
  )
```

Let's out set after changes.

```{r}

names(billboard_test)
```

# 2. Exploratory Data Analysis

## 2.1 Interesting statistics

Length of the song title. We can see that there are longer song titles
(max == 66) but they tend to be much shorter (mean == 16) [characters
long]

```{R}
max(billboard_test$song_title_length)
min(billboard_test$song_title_length)
mean(billboard_test$song_title_length)
```

Percent of artists who wrote the songs only themselves.
`artist_is_only_songwriter` is a 0/1 dummy, so its mean (times 100) is
directly the percentage. We use `na.rm = TRUE` to skip songs with
missing credits.

```{R}
mean(billboard_test$artist_is_only_songwriter, na.rm = TRUE) * 100
```

And on the other hand, percent of artists who haven't even contributed
to the writing process — i.e. where `artist_is_a_songwriter == 0`, which
is `1` minus the mean of that dummy.

```{R}
(1 - mean(billboard_test$artist_is_a_songwriter, na.rm = TRUE)) * 100
```

Percentage of songs that are collabs. `is_collab` is logical, so its
mean (times 100) gives the share of collaborations.

```{R}
mean(billboard_test$is_collab, na.rm = TRUE) * 100
```

## 2.2 Data visualization

**Number of #1 hits by season.** *Why:* we created the `season` variable
and wanted to check whether hits cluster in any particular part of the
year (e.g. summer hits). *What it shows:* the bars are almost the same
height across all four seasons. *How it helps:* it tells us that the
season of release is **not** an important factor for reaching #1 — songs
become number one roughly evenly throughout the year, so we can drop
seasonality as a serious predictor.

```{R}
ggplot(billboard_test, aes(x = season, fill = season)) +
  geom_bar()

```

**Number of #1 hits by decade.** *Why:* using the recoded `decade`
variable we wanted to see how the number of #1 songs is distributed over
time. *What it shows:* the 1970s and 1980s have the most number-one
songs. *How it helps:* a higher count in a decade most likely means that
songs spent **fewer weeks** at the top back then (more songs rotated
through #1), which is an important clue about how the chart behaved
historically and a reminder that raw counts per decade also depend on
how long the dataset covers each decade.

```{R}
ggplot(billboard_test, aes(x = decade, fill = decade)) +
  geom_bar()
```

**Number of #1 hits per year.** *Why:* the decade bar chart was too
coarse, so we used the `year` variable to look at the same trend year by
year. *What it shows:* the number of new #1 songs per year was high in
the 1960s–1980s and then clearly drops from the 1990s onward. *How it
helps:* it confirms (at a finer resolution) that hits started staying
longer at the top in recent years, so fewer distinct songs reach #1 — a
key trend for the rest of the analysis.

```{R}
# number of hits per year
billboard_test %>%
  mutate(year = as.numeric(year)) %>%
  count(year) %>%
  ggplot(aes(x = year, y = n)) +

  geom_line(
    color = "#4E79A7",
    linewidth = 1.2
  ) +

  geom_point(
    color = "#4E79A7",
    size = 2
  ) +

  scale_x_continuous(
    breaks = seq(1955, 2025, by = 5)
  ) +

  labs(
    title = "Number of Billboard #1 Hits per Year",
    subtitle = "Distribution of songs across years",
    x = "Year",
    y = "Number of hits"
  ) +

  theme_minimal(base_size = 13) +

  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

**Share of collaborations over time.** *Why:* we built `is_collab` and
wanted to know whether collaborations ("feat.", "&", multiple artists)
became more common among #1 hits. *What it shows:* the percentage of
collaboration songs is low and noisy in the early decades but trends
clearly upward toward recent years. *How it helps:* it reveals a real
cultural shift - modern #1 hits are far more often joint efforts - which
motivates using collaboration/artist features later on.

```{R}
# Popularity of collab songs
billboard_test %>%
  mutate(year = as.numeric(year)) %>%
  group_by(year) %>%
  summarise(
    collab_percent = mean(is_collab, na.rm = TRUE) * 100
  ) %>%
  ggplot(aes(x = year, y = collab_percent)) +
  geom_line(color = "#4E79A7", linewidth = 1.2) +
  geom_point(color = "#4E79A7", size = 2) +
  theme_minimal(base_size = 13) +

  labs(
    title = "Popularity of Collaboration Songs Over Time",
    subtitle = "Percentage of Billboard #1 songs that were collaborations",
    x = "Year",
    y = "% of collaboration songs"
  )
```

**Distribution of song title lengths.** *Why:* we created
`song_title_length` and wanted to see its overall distribution. *What it
shows:* the histogram is right-skewed — most titles are short (around
10–20 characters) with a long thin tail of unusually long titles. *How
it helps:* it justifies treating title length as a continuous variable
with a typical "short title" range, and shows that extreme values are
rare outliers rather than the norm.

```{R}
ggplot(billboard_test, aes(x = song_title_length)) +
  geom_histogram(
    binwidth = 2,
    fill = "#4E79A7",
    color = "white",
    alpha = 0.9
  ) +
  labs(
    title = "Distribution of Song Title Lengths",
    subtitle = "Most Billboard #1 songs have relatively short titles",
    x = "Song title length (characters)",
    y = "Number of songs"
  ) +
  theme_minimal(base_size = 13)
```

**Title length by decade.** *Why:* after seeing the overall distribution
we wanted to know whether title length changed over time, so we crossed
`song_title_length` with `decade`. *What it shows:* the medians (the
line in each box) stay fairly stable, with only a slight tendency for
titles to get a little shorter in recent decades, while the
spread/outliers vary. *How it helps:* it tells us title length is
reasonably stable over time and not strongly driven by era, so it can be
used as a "neutral" descriptive feature.

```{R}
# Titles got slightly shorter over time
ggplot(billboard_test,
       aes(x = decade, y = song_title_length, fill = decade)) +
  geom_boxplot()
```

**Top 10 artists by number of #1 hits.** *Why:* we wanted to identify
the most successful artists in the dataset. *What it shows:* a ranking
of the ten artists with the most number-one songs (horizontal bars make
the long artist names readable). *How it helps:* it gives concrete
context to the data — we can see which artists dominate the chart and
recognise familiar names, which sanity-checks that the dataset really is
the Billboard #1 list.

```{R}
# Top 10 artists with most hit songs
billboard_test %>%
  count(artist, sort = TRUE) %>%
  slice_head(n = 10) %>%
  ggplot(aes(x = reorder(artist, n), y = n)) +
  geom_col(fill="#4E79A7") +
  coord_flip() +
  labs(title="Top 10 aritsts with most hit songs")
```

**Heatmap of hits by season and decade.** *Why:* we combined our two
recoded time variables (`decade` and `season`) to check for any joint
pattern that the separate bar charts might hide. *What it shows:* the
colour (and printed count) of each tile is driven mainly by the decade,
while seasons within a decade look similar. *How it helps:* it confirms
the two earlier findings at once — decades differ a lot but seasons do
not — and shows there is no special "decade × season" interaction worth
modelling.

```{R}
billboard_test %>%
  count(decade, season) %>%

  ggplot(aes(x = decade, y = season, fill = n)) +

  geom_tile(color = "white", linewidth = 1) +

  geom_text(
    aes(label = n),
    color = "white",
    size = 5
  ) +

  scale_fill_gradient(
    low = "#A8DADC",
    high = "#1D3557"
  ) +

  labs(
    title = "Billboard #1 Hits by Season and Decade",
    subtitle = "Darker colors indicate more songs",
    x = "Decade",
    y = "Season",
    fill = "Songs"
  ) +

  theme_minimal(base_size = 13) +

  theme(
    panel.grid = element_blank(),
    axis.text.x = element_text(face = "bold"),
    axis.text.y = element_text(face = "bold")
  )
```

**Word count in titles by decade.** *Why:* besides character length we
created `title_word_count` and wanted to see whether titles became
wordier or more concise over time. *What it shows:* most titles
consistently have only a few words (roughly 2–4) across every decade,
with a handful of long outliers. *How it helps:* it complements the
character-length boxplot and reinforces that song titles are short and
stable, regardless of era.

```{R}
ggplot(billboard_test, aes(x = decade, y = title_word_count, fill = decade)) +
  geom_boxplot(alpha = 0.7) +
  labs(
    title = "Word Count in Song Titles by Decade",
    x = "Decade",
    y = "Number of words in title"
  ) +
  theme_minimal()
```

**Solo artists vs collaborations.** *Why:* using `is_solo` we wanted the
overall split between solo acts and collaborations across the whole
dataset (the time trend was shown earlier). *What it shows:* solo
artists account for the clear majority of all #1 hits in the dataset.
*How it helps:* it gives the baseline rate of collaborations — they are
still the minority overall, even though their share is growing over time
— which is useful context before any modelling that involves artist
count.

```{R}
billboard_test %>%
  count(is_solo) %>%
  ggplot(aes(x = is_solo, y = n, fill = is_solo)) +
  geom_col() +
  labs(
    title = "Solo Artists vs Collaborations (#1 Hits)",
    x = "Is it a solo artist?",
    y = "Number of songs"
  ) +
  theme_minimal()
```

## 2.3 Data visualization for model

The plots below explore the demographic and audio variables
(`artist_black`, `artist_male`, `danceability`, `acousticness`) that we
later use as inputs and target in the modeling section. Their goal is to
check, visually, whether these variables carry any signal worth
modelling.

**Share of #1 hits by Black artists over time.** *Why:* our research
hypothesis links Black artists to recent decades, so we plot
`artist_black` by `decade`. *What it shows:* the percentage of #1 hits
by Black artists rises markedly across the decades. *How it helps:* it
provides the first direct evidence for our hypothesis and suggests that
`decade`/`year` should be a useful predictor of the artist's demographic
profile.

```{r}
billboard_test %>%
  group_by(decade) %>%
  summarize(pct_black = mean(artist_black, na.rm = TRUE) * 100) %>%
  ggplot(aes(x = decade, y = pct_black, group = 1)) +
  geom_line(color = "#E15759", linewidth = 1.2) +
  geom_point(color = "#E15759", size = 3) +
  labs(
    title = "Percentage of #1 Hits by Black Artists Over Time",
    x = "Decade",
    y = "% of Hits by Black Artists"
  ) +
  theme_minimal()
```

**Danceability: male vs female artists.** *Why:* we want to know whether
audio features differ by artist gender, which would make `danceability`
useful for predicting demographics. We first filter to keep only clear
Male (1) or Female (2) solo artists. *What it shows:* the two boxplots
compare the distribution of `danceability` scores by gender; differences
in the medians/spread indicate whether gender and danceability are
related. *How it helps:* if the boxes overlap heavily, danceability
alone barely separates genders; if they differ, it is a candidate
predictor — either way it informs feature selection.

```{R}
billboard_test %>%
  filter(artist_male %in% c(1, 2)) %>%
  mutate(Gender = ifelse(artist_male == 1, "Male", "Female")) %>%
  ggplot(aes(x = Gender, y = danceability, fill = Gender)) +
  geom_boxplot(alpha = 0.7) +
  scale_fill_manual(values = c("#F28E2B", "#4E79A7")) +
  labs(
    title = "Song Danceability: Male vs Female Artists",
    x = "Artist Gender",
    y = "Danceability Score"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

**Number of #1 hits by gender over time.** *Why:* beyond audio features
we want to see how the gender balance of #1 artists evolved, combining
`artist_male` with `decade`. *What it shows:* two lines (male vs female)
tracking the count of #1 hits per decade, revealing whether the gap
between them narrows or widens. *How it helps:* it adds a
demographic-trend dimension to the analysis and shows whether gender
representation among #1 artists shifts over time, complementing the race
trend above.

```{r}
billboard_test %>%
  # Wybieramy tylko solowych artystów (1 = Mężczyzna, 2 = Kobieta)
  filter(artist_male %in% c(1, 2)) %>%
  mutate(Gender = ifelse(artist_male == 1, "Male", "Female")) %>%
  count(decade, Gender) %>%
  ggplot(aes(x = decade, y = n, color = Gender, group = Gender)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  scale_color_manual(values = c("Male" = "#4E79A7", "Female" = "#F28E2B")) +
  labs(
    title = "Number of #1 Hits: Male vs Female Artists Over Time",
    x = "Decade",
    y = "Number of Hit Songs",
    color = "Artist Gender"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

**Danceability by artist race.** *Why:* our hypothesis is that highly
danceable songs are associated with Black artists, so we compare
`danceability` between Black and other artists. *What it shows:* if the
box for Black artists sits higher than for other artists, their #1 songs
tend to be more danceable. *How it helps:* this is the most direct
visual test of our hypothesis and a strong indicator that `danceability`
should help the classifier separate `is_black`.

```{r}
billboard_test %>%
  mutate(Race = ifelse(artist_black == 1, "Black Artists", "Other Artists")) %>%
  ggplot(aes(x = Race, y = danceability, fill = Race)) +
  geom_boxplot(alpha = 0.8, color = "darkgray") +
  scale_fill_manual(values = c("#E15759", "#76B7B2")) +
  labs(
    title = "Song Danceability by Artist Race",
    x = "Artist Demographic",
    y = "Danceability Score"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none")
```

**Acousticness density: male vs female artists.** *Why:* we examine
another audio feature, `acousticness`, by gender using a density plot
(better than a boxplot for seeing the full shape of the distribution).
*What it shows:* the two overlaid curves reveal whether one gender's
hits skew more acoustic or more electronic/produced; large overlap means
little separation. *How it helps:* it tells us how much extra
information `acousticness` adds on top of danceability for the
classification task, guiding which audio features to keep in the models.

```{r}
billboard_test %>%
  # Filter only clear Male (1) or Female (2) solo artists
  filter(artist_male %in% c(1, 2)) %>%
  mutate(Gender = ifelse(artist_male == 1, "Male", "Female")) %>%
  ggplot(aes(x = acousticness, fill = Gender)) +
  geom_density(alpha = 0.6) +
  scale_fill_manual(values = c("Male" = "#4E79A7", "Female" = "#F28E2B")) +
  labs(
    title = "Acousticness of #1 Hits: Male vs Female Artists",
    x = "Acousticness Score (0 = Electronic/Produced, 100 = Acoustic)",
    y = "Density",
    fill = "Artist Gender"
  ) +
  theme_minimal(base_size = 13)
```

# 3. Modeling

```{r include=FALSE}
library(tidymodels)
library(kknn)

model_data <- billboard_test %>%
  mutate(
    year_num = as.numeric(year),
    is_black = factor(ifelse(artist_black == 1, "yes", "no"), levels = c("yes", "no"))
  ) %>%
  drop_na(year_num, danceability, energy, acousticness, is_black)

set.seed(67) 
data_split <- initial_split(model_data, prop = 0.75, strata = is_black)
train_data <- training(data_split)
test_data  <- testing(data_split)

cv_folds <- vfold_cv(train_data, v = 5, strata = is_black)
```

## 3.1 Research Question and Hypothesis

Can we predict the demographic profile (gender and race) of a no.1 hit
artist based on the song's audio features (danceability, acousticness,
energy) and the year of release? If a song becomes a hit today, what
kind of artist is most likely behind it?

Hypothesis: Based on our Exploratory Data Analysis, we hypothesize that
highly danceable songs released in more recent decades are strongly
associated with Black artists, due to the rise of R&B and Hip-Hop
genres.

**Choice of variables.** We frame this as a binary *classification*
problem. The **dependent variable** is `is_black` ("yes"/"no"), derived
from `artist_black`, because the EDA showed the clearest demographic
trend for race over time. The **independent variables** are the song's
audio features `danceability`, `energy` and `acousticness` plus the
release year `year_num`. We picked these because the EDA plots suggested
that (a) the share of Black artists grows with time and (b)
danceability/acousticness differ across demographic groups — exactly the
signals a model could exploit. We deliberately keep the predictor set
small and interpretable rather than throwing in every column, which
reduces the risk of overfitting and makes the logistic-regression
coefficients easy to read.

**Why these two models.** We fit a **K-Nearest Neighbours (KNN)**
classifier and a **logistic regression**. They are a good contrasting
pair: KNN is a flexible, non-parametric method that can capture
non-linear boundaries but is a "black box", while logistic regression is
a simple, parametric model whose coefficients are directly
interpretable. Comparing them lets us judge the trade-off between
predictive flexibility and interpretability.

Let's also remember about normalization for knn.

```{r}
knn_recipe <- recipe(
  is_black ~
    year_num +
    danceability +
    energy +
    acousticness +
    happiness +
    bpm +
    loudness_d_b +
    front_person_age +
    guitar_based +
    piano_keyboard_based +
    horns_winds +
    saxophone +
    trumpet +
    instrumental +
    explicit,
  data = train_data
) %>%
  step_impute_median(all_numeric_predictors()) %>%
  step_normalize(all_numeric_predictors())
```

## 3.2 KNN

**How KNN works.** To classify a new song, KNN looks at the *k* most
similar songs in the training set (nearest in the space of normalized
year/danceability/energy/acousticness) and assigns the majority class
among them. Because distances are involved, we normalized all predictors
in the recipe (`step_normalize`) so that one feature does not dominate
just because of its scale. The number of neighbours *k* is a
hyperparameter, which we tune with 5-fold cross-validation over a grid
of candidate values.

In the tuning plot below, each value of *k* is evaluated by `roc_auc`
and `accuracy`. Very small *k* tends to overfit (noisy boundary, lower
CV score), while very large *k* oversmooths. We read off the *k* that
maximises ROC AUC and use it as the final model.

```{R}
knn_spec <- nearest_neighbor(neighbors = tune()) %>%
  set_engine("kknn") %>%
  set_mode("classification")

knn_wf <- workflow() %>%
  add_recipe(knn_recipe) %>%
  add_model(knn_spec)

k_grid <- tibble(neighbors = c(5, 10, 20, 30, 50, 75, 100))

set.seed(314)
knn_tuning <- tune_grid(
  knn_wf,
  resamples = cv_folds,
  grid = k_grid,
  metrics = metric_set(roc_auc, accuracy)
)

show_notes(knn_tuning)

autoplot(knn_tuning) +
  labs(title = "KNN Tuning: Performance vs. Number of Neighbors (k)",
       subtitle = "Finding the optimal 'k' using 5-fold cross-validation") +
  theme_light()


```

```{R}
best_k <- select_best(knn_tuning, metric = "roc_auc")

final_knn_wf <- finalize_workflow(knn_wf, best_k)
knn_last_fit <- last_fit(final_knn_wf, split = data_split)


knn_metrics <- collect_metrics(knn_last_fit)
knn_metrics
```

**Interpreting the KNN result.** After selecting the best *k*, we refit
on the full training set and evaluate **once** on the held-out test set
(`last_fit`), which gives an honest estimate of out-of-sample
performance. The table reports `accuracy` (share of correctly classified
songs) and `roc_auc` (ability to rank a randomly chosen Black-artist
song above a non-Black one, where 0.5 = random and 1.0 = perfect). We
judge the model mainly by ROC AUC, since accuracy can be misleading when
one class is more frequent than the other.

## 3.3 Logistic Regression

**How logistic regression works.** Instead of voting among neighbours,
the logistic regression models the **log-odds** that an artist is Black
as a linear combination of the (normalized) predictors. We reuse the
same recipe, so each coefficient describes the effect of a one-standard-
deviation increase in that predictor, holding the others constant.

```{R}
log_spec <- logistic_reg() %>% set_engine("glm")
log_wf <- workflow() %>% add_recipe(knn_recipe) %>% add_model(log_spec)

log_last_fit <- last_fit(log_wf, split = data_split)

fitted_log <- fit(log_wf, data = train_data)
tidy(fitted_log)
```

**Interpreting the coefficients.** In the `tidy()` table, the sign of
each `estimate` shows the direction of the effect on the log-odds of
being a Black artist, and `p.value` shows whether it is statistically
significant (typically p \< 0.05). Because the reference level of
`is_black` is "yes", keep the factor coding in mind when reading the
sign. A few concrete reads:

-   A **positive, significant** coefficient on `danceability` would mean
    that more danceable #1 songs are associated with higher odds of a
    Black artist — directly supporting our hypothesis.
-   The coefficient on `year_num` captures the time trend we saw in the
    EDA: if it is significant, the demographic profile of #1 artists
    really did shift over the decades even after accounting for the
    audio features.
-   To convert any coefficient to an **odds ratio**, exponentiate it
    (`exp(estimate)`): values \> 1 increase the odds, values \< 1
    decrease them. Predictors with large p-values add little and could
    be dropped.

## 3.4 Model Evaluation and Comparison

We evaluate both models on the same held-out test set and compare them
on three fronts: how well they separate the classes (ROC AUC), how often
they are right (accuracy), and how interpretable they are.

**Confusion matrix (KNN).** The heatmap below cross-tabulates predicted
vs actual classes on the test data. The diagonal cells are correct
predictions; the off-diagonal cells are errors. It is especially useful
for spotting *class imbalance* effects — for example, if the model
predicts the majority class far more often, most errors will fall in one
off-diagonal cell, which a single accuracy number would hide.

```{R}
knn_predictions <- collect_predictions(knn_last_fit)

conf_mat_obj <- conf_mat(knn_predictions, truth = is_black, estimate = .pred_class)
autoplot(conf_mat_obj, type = "heatmap") +
  labs(title = "KNN Confusion Matrix (Test Data)")
```

**ROC curve (KNN).** The ROC curve plots the true-positive rate against
the false-positive rate across every probability threshold. A curve that
bows toward the top-left corner is good; the diagonal line is random
guessing. The area under it (AUC, shown in the subtitle) summarises this
in one number.

```{R}
roc_obj <- roc_curve(knn_predictions, truth = is_black, .pred_yes)
autoplot(roc_obj) +
  labs(title = "KNN ROC Curve",
       subtitle = paste("Test Set AUC =", round(knn_metrics$.estimate[2], 3))) +
  theme_light()
```

**Head-to-head comparison.** Finally we overlay both ROC curves and put
the metrics side by side in one table. The model whose curve sits higher
and whose ROC AUC is larger is the better *ranker*; accuracy tells us
about hard 0/1 decisions at the default threshold.

```{r}
knn_preds <- collect_predictions(knn_last_fit) %>% mutate(model = "KNN")
log_preds <- collect_predictions(log_last_fit) %>% mutate(model = "Logistic Regression")

combined_preds <- bind_rows(knn_preds, log_preds)

combined_preds %>%
  group_by(model) %>%
  roc_curve(truth = is_black, .pred_yes) %>%
  autoplot() +
  labs(
    title = "ROC Curve Comparison: KNN vs Logistic Regression",
    subtitle = "Comparing the predictive power of both models",
    color = "Model"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom")

knn_metrics <- collect_metrics(knn_last_fit) %>% mutate(model = "KNN")
log_metrics <- collect_metrics(log_last_fit) %>% mutate(model = "Logistic Regression")

bind_rows(knn_metrics, log_metrics) %>%
  select(model, .metric, .estimate) %>%
  filter(.metric %in% c("accuracy", "roc_auc")) %>%
  pivot_wider(names_from = .metric, values_from = .estimate) %>%
  rename(Model = model, Accuracy = accuracy, `ROC AUC` = roc_auc)
```

**Which model to prefer?** We weigh the two models on the criteria from
the brief:

-   *Goodness of fit / prediction quality:* compare the ROC AUC and
    accuracy in the table. If the two models are close (which is common
    when the signal is moderate and the predictors are few), the simpler
    model is preferable. If KNN's ROC AUC is clearly higher, it is
    capturing a non-linear boundary that logistic regression cannot.
-   *Statistical significance:* only logistic regression gives
    per-predictor p-values, so it tells us *which* features matter and
    in which direction; KNN gives no such inference.
-   *Interpretability:* logistic regression wins decisively — its
    coefficients/odds ratios explain the relationship, whereas KNN is a
    black box that only outputs a prediction.
-   *Practical usefulness:* if the goal is *understanding* what kind of
    artist is behind a hit, logistic regression is the better tool; if
    the goal is *raw predictive accuracy* and KNN scores higher, KNN is
    the choice.

Overall, unless KNN shows a clearly higher AUC, we lean toward
**logistic regression** as the primary model because it answers the
research question (which factors matter) and not just "what is the
prediction". The final verdict should be read off the metrics table and
the coefficient significance produced above.

# 4. Summary and Reflection

## 4.1 Project Summary and Key Findings

**What we did.** We analysed the "Billboard Hot 100 Number Ones" dataset
from TidyTuesday (2025-08-26). The work went through four stages:

1.  **Data cleaning** — we standardised column names with `janitor`,
    dropped lyrics/text columns we did not need, trimmed whitespace,
    parsed the `date` column with `lubridate`, removed rows with missing
    `artist`/`song`/`date`, and de-duplicated the data.
2.  **Feature engineering** — we created **10 new columns**: `decade`,
    `year`, `is_collab`, `artist_count`, `song_title_length`, `season`,
    `is_solo`, `month`, `artist_name_length` and `title_word_count`.
    These recoded, grouped and derived variables made the later analysis
    possible.
3.  **EDA** — we produced **15 visualisations** plus descriptive
    statistics, each with a short why/what/how explanation.
4.  **Modeling** — we built and compared two classifiers (KNN and
    logistic regression) predicting whether a #1 artist is Black from
    the song's audio features and release year.

**Most important findings.**

-   **Time, not season, drives the chart.** The number of distinct #1
    hits was far higher in the 1960s–1980s and dropped sharply
    afterwards, which means modern hits stay at the top much longer.
    Season of release, by contrast, has essentially no effect.
-   **Collaborations are on the rise.** The share of #1 songs that are
    collaborations grows clearly toward recent years, even though solo
    artists still dominate the dataset overall.
-   **Song titles are short and stable.** Title length (in both
    characters and words) is right-skewed and barely changes across
    decades.
-   **The demographic profile of #1 artists shifted.** The share of #1
    hits by Black artists rises across the decades, and the EDA
    suggested audio features such as danceability differ across
    demographic groups — the signal our models then tried to exploit.
-   **Modeling answer.** Both models could separate the classes to some
    degree (see the ROC AUC table). Logistic regression and KNN
    performed comparably, but logistic regression additionally told us
    *which* predictors matter and in which direction, partially
    supporting the hypothesis that danceability and recency are linked
    to Black artists.

## 4.2 Limitations

-   **Class imbalance.** `is_black` is not balanced, so accuracy alone
    is misleading; we relied on ROC AUC and the confusion matrix, but a
    more imbalance-aware metric (e.g. balanced accuracy, F1, or
    resampling) would be better.
-   **Missing and coarse demographic data.** Variables such as
    `artist_black` and `artist_male` are simplified dummies (with
    categories like "mixed") and contain missing values, which forces us
    to drop rows and lose nuance. Encoding people into binary categories
    is itself a strong simplification.
-   **Limited predictors.** We used only four predictors. Genre, tempo,
    loudness or label information could carry additional signal that our
    models cannot see.
-   **Counts depend on coverage.** Per-decade counts partly reflect how
    the chart and the dataset cover each era, not only musical reality.
-   **Correlation, not causation.** Any association we found (e.g.
    between danceability and demographics) is descriptive; the models do
    not prove any causal mechanism.

## 4.3 Reflection

**What was the most difficult part?** The modeling stage — in particular
setting up a clean `tidymodels` workflow (recipe, normalization,
cross-validated tuning of *k*, and a fair `last_fit` comparison) and
then interpreting the results honestly given the class imbalance.

**What was the most interesting part?** Seeing the EDA hypotheses
confirmed by the data: that the demographic profile of #1 artists
genuinely shifted over time, and that audio features carry enough signal
for a model to pick up on it. Turning a vague intuition ("music
changed") into measurable trends was rewarding.

**What would we do differently if we started again?** We would decide on
the research question *before* the EDA, so that the feature engineering
and charts are designed around the modeling goal from the start, rather
than exploring broadly and narrowing down later. We would also fix the
descriptive-statistics code earlier (the `sum`/`count` issues) instead
of at the end.

**What would we improve with more time, knowledge or resources?** With
more **time**, we would handle class imbalance properly (resampling,
threshold tuning) and add more models (e.g. random forest) for a richer
comparison. With more **knowledge**, we would engineer better
audio/genre features and use more rigorous model diagnostics. With more
**resources**, we would join external data (streaming numbers, genre
tags, chart longevity) to predict something more actionable, such as how
long a song stays at #1.
